Reads, QC and trimming

### 2024-04-03 episodes/blast.Rmd
Removing an extra closing fence
Finalizing Blast, upgrading MSA
Fixing a few issues with Mobaxterm on Windows and not waiting for the blast to complete.

::: instructors

@@

```bash
interactive -A uppmax2024-2-10 -M snowy -t 4:00:00 --reservation=uppmax2023-2-10_x
```

@@

@@

@@

`refseq_select_prot` and `refseq_protein` are good candidates. The former is smaller than the latter. Note that these two runs will take time, so run only the one to `refseq_select_prot`

```bash
blastp -db refseq_select_prot -query rpoB_ecoli.fasta -evalue 1e-6 -outfmt 6 > rpoB_refseq_select.tab
```

Now count the rows to have the number of hits
```bash
wc -l rpoB_refseq_select.tab
```

@@

```bash
psiblast -help
```

>>>>

::: instructor

@@

```bash
interactive -A uppmax2024-2-10 -M snowy -t 4:00:00 
```

@@

To bring up a local terminal on your Windows computer, click on the "+" sign on the main window of MobaXTerm. If that doesn't work, use SFTP. Click on Session, SFTP, and fill in the Remote host as usual `rackham.uppmax.uu.se` and your username. Navigate to `/proj/g2020004/nobackup/3MK013/<username>/blast` on the right panel, then drag and drop files between your computer and Uppmax.

@@

`refseq_select_prot` and `refseq_protein` are good candidates. The former is smaller than the latter. Note that these two runs will take time, so run only the one to `refseq_select_prot`. We will run it in the background (adding a `&` at the end of the command) so that you can continue with other tasks and come back to that one later.

@@

```bash
blastp -db refseq_select_prot -query rpoB_ecoli.fasta -evalue 1e-6 -outfmt 6 > rpoB_refseq_select.tab &
```

You can check whether `blastp` is still running by typing `ps`:

```bash
ps
```

```output
  PID TTY          TIME CMD
13920 pts/64   00:00:00 bash
33117 pts/64   00:00:00 blastp
35427 pts/64   00:00:00 ps
```

If you see `blastp` there, it means it is still running.

When it has finished (it may take up to 30 minutes), count the rows to have the number of hits:

```bash
wc -l rpoB_refseq_select.tab
```

@@

```bash
psiblast -help
```
::::::::

### 2024-04-03 episodes/quality-control.Rmd
Fixing a few issues with Mobaxterm on Windows and not waiting for the blast to complete.

---
title: Assessing Read Quality and Trimming
teaching: 30
exercises: 20
---

Let's check if the downloads finished. Let's have a look at the files.

::::::::::::::::::::::::::::::::::::::: objectives

- Interpret a FastQC plot summarizing per-base quality across all reads.
- Clean FastQC reads for further analysis.
- Use `for` loops to automate operations on multiple files.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I describe the quality of my data?
- How can I get rid of sequence data that doesn't meet my quality standards?

::::::::::::::::::::::::::::::::::::::::::::::::::

```source
$ ls
```

```output
ERR029206_1.fastq.gz
ERR029206_2.fastq.gz 
```

All new files start with ERR and have a `fastq.gz` extension. It means they are in fastq format and compressed. Let's make a copy of one file and uncompress this file to inspect it further

```bash
cp ERR029206_1.fastq.gz ERR029206_1.copy.fastq.gz
gunzip ERR029206_1.copy.fastq.gz
```

@@

![](fig/Workflow.png){alt='Schematic workflow of the bioinformatics workflow'}

@@

![](fig/var_calling_workflow_qc.png){alt='Quality control workflow'}

@@

| Line   | Description                                                                                                  | 
| ------ | ------------------------------------------------------------------------------------------------------------ |
| 1      | Always begins with '@' and then information about the read                                                   | 
| 2      | The actual DNA sequence                                                                                      | 
| 3      | Always begins with a '+' and sometimes the same info in line 1                                               | 
| 4      | Has a string of characters which represent the quality scores; must have same number of characters as line 2 |

We can view the first complete read in one of the files our dataset by using `head` to look at the first four lines.

```bash
$ head -n4 ERR029206_1.copy.fastq
```

@@

We can now see that the quality of an `N` (`'`) is 6, which corresponds to a over 25% error probability ($10^{-\frac{6}{10}} \cong 0.25 $)

> ### Quality Encodings Vary
> 
> Although we've used a particular quality encoding system to demonstrate interpretation of
> read quality, different sequencing machines use different encoding systems. This means that,
> depending on which sequencer you use to generate your data, a `#` may not be an indicator of
> a poor quality base call.

:::::::::::::::::::::::::::::::::::::::::  callout

This mainly relates to older Solexa/Illumina data,
but it's essential that you know which sequencing platform was
used to generate your data, so that you can tell your quality control program which encoding
to use. If you choose the wrong encoding, you run the risk of throwing away good reads or
(even worse) not throwing away bad reads!

::::::::::::::::::::::::::::::::::::::::::::::::::

@@

![](fig/good_quality1.8.png){alt='A FastQC output showing reads with good quality'}

@@

![](fig/bad_quality1.8.png){alt='A FastQC output showing reads with bad quality'}

@@

```bash
$ cd ~/molepi/data/
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Exercise

We downloaded the samples. How many FASTQ files are in this dataset? Why is this? What is the difference between the two files that start with the same name? Discuss with your neighbour.

:::::::::::::::  solution

### Solution

There are 2 FASTQ files. Each sample has [paired-end reads](https://thesequencingcenter.com/knowledge-base/what-are-paired-end-reads/).

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

FastQC can accept multiple file names as input, including files that are compressed (`.gz`) so we can use the `*.fastq.gz` wildcard to run FastQC on all of the compressed FASTQ files in this directory.

```bash
$ fastqc *.fastq.gz
```

@@

```bash
$ ls
```

```output
...
ERR029206_1.copy.fastq
ERR029206_1.fastq.gz
ERR029206_1_fastqc.html
ERR029206_1_fastqc.zip
ERR029206_2.fastq.gz
ERR029206_2_fastqc.html
ERR029206_2_fastqc.zip
```

@@

If we are working on our local computers, we may display each of these HTML files as a webpage:

```bash
$ firefox ERR029206_1_fastqc.html
```

@@

:::::::::::::::::::::::::::::::::::::::  challenge

### Exercise

Discuss your results with a neighbor. Do the sample looks good in terms of per base sequence quality?

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

### Exercise

What are the read lengths of the different samples?

::::::::::::::::::::::::::::::::::::::::::::::::::

@@

```bash
$ seqtk
```

@@

| option  | meaning                                           | 
| ------  | --------------------------------------------------|
| `seq`   | common transformation of FASTA/Q                  | 
| `comp`  | get the nucleotide composition of FASTA/Q         | 
| `trimfq`| trim FASTQ using the Phred algorithm              |

@@

```bash
$ seqtk trimfq
```

| step   | meaning                                                        | 
| ------ | ---------------------------------------------------------------|
| `-q`   | error rate threshold (disabled by -b/-e) [0.05]                | 
| `-l`   | maximally trim down to INT bp (disabled by -b/-e) [30]         | 
| `-b`   | trim INT bp from left (non-zero to disable -q/-l) [0]          | 
| `-e`   | trim INT bp from right (non-zero to disable -q/-l) [0].        |

@@

```bash
$ seqtk trimfq -q 0.01 ERR01_1.fastq.gz > ERR01_1_trim.fastq
```

@@

```bash
$ cd ~/molepi/data/
```

@@

```bash
$ seqtk trimfq -q 0.01 ERR029206_1.fastq.gz > ERR029206_1_trim.fastq
```

@@

:::::::::::::::::::::::::::::::::::::::  challenge

### Exercise

@@

:::::::::::::::  solution

### Solution

```bash
seqtk fqchk ERR029206_1.copy.fastq | head -n 3
```

@@

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

@@

```bash
$ for infile in *.fastq.gz
do
outfile="${infile}"_trim.fastq
seqtk trimfq -q 0.01 "${infile}" >  "${outfile}" 
done
```

@@

```bash
$ ls
```

```output
...
ERR029206_1.copy.fastq 
ERR029206_1.fastq.gz
ERR029206_1.fastq.gz_trim.fastq 
ERR029206_1_trim.fastq 
ERR029206_2.fastq.gz
ERR029206_2.fastq.gz_trim.fastq
GCF_000195955.2_ASM19595v2_genomic.fna
```

@@

```bash
$ mkdir trimmed_fastq
$ mv *fastq.gz_trim* trimmed_fastq
$ cd trimmed_fastq
$ ls
```

```output
...
ERR029206_1.fastq.gz_trim.fastq 
ERR029206_2.fastq.gz_trim.fastq 
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

@@

:::::::::::::::  solution

### Solution

@@

Coverage =  #bases (forward + reverse) / genome size
In this case: 
125.18 = (287211920 + 265048852) / 4411532

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: keypoints

- In bioinformatic workflows the output of one tool is the input of the other.
- FastQC is used to judge the quality of sequencing reads.
- Data cleaning is an essential step in a genomics pipeline.

::::::::::::::::::::::::::::::::::::::::::::::::::

>>>>

---
title: "Reads, QC and trimming"
teaching: 0
exercises: 240
---

This episode starts a series of labs where you will trace an outbreak of *Mycobacterium tuberculosis*, from downloading the reads to making assumptions about the transmissions between patients.

The first part, here, describes how to retrieve all necessary data from public repositories - the raw sequenced data of our isolates and a reference genome. It also introduces for loops which we will use for the rest of our analysis.

::: objectives

- How do you use genomics and phylogenetics to trace a microbial outbreak?
- Create a file system for a bioinformatics project.
- Download files necessary for further analysis.
- Use 'for' loops to automate operations on multiple files
- Interpret a FastQC plot summarizing per-base quality across all reads.
- Clean FastQC reads for further analysis.
- Use `for` loops to automate operations on multiple files.

:::

::: questions

- Trace an outbreak of Mycobacterium, from reads to phylogenetic tree.
- How can I organize my file system for a new bioinformatics project?
- How and where can data be downloaded?
- How can I describe the quality of my data?
- How can I get rid of sequence data that doesn't meet my quality standards?

:::

## Introduction

This part of the course is largely inspired from [another tutorial](https://github.com/aschuerch/MolecularEpidemiology_AnalysisWGS) developed by [Anita Schürch](https://aschuerch.github.io/). We will use (part of) the data described in [Bryant et al., 2013](https://bmcinfectdis.biomedcentral.com/articles/10.1186/1471-2334-13-110).

## Background of the data

Epidemiological contact tracing through interviewing of patients can identify potential chains of patients that transmitted an infectious disease to each other. Contact tracing was performed following the [stone-in-the-pond principle](https://www.ncbi.nlm.nih.gov/pubmed/1643300), which  interviews and tests potential contacts in concentric circles around a potential bash case.

Tuberculosis (TB) is an infectious disease caused by *Mycobacterium tuberculosis*. It mostly affects the lungs. An infection with *M. tuberculosis* is often asymptomatic (latent infection). Only in about 10% of the cases the latent infection progresses to an active infection during a patients lifetime, which, if untreated, leads to death in about half of the cases. The symptoms of an active TB infection include cough, fever, night sweats, weight loss etc. An active TB infection can spread. Once exposed, people often have latent TB. To identify people with latent TB, a [skin test](https://www.cdc.gov/tb/publications/factsheets/testing/skintesting.htm) can be applied.

Here we have 7 tuberculosis patients with active TB, that form three separate clusters of potential transmission as determined by epidemiological interviews. Patients were asked if they have been in direct contact with each other, or if they visited the same localities. From all patients, a bacterial isolate was grown, DNA isolated, and whole-genome sequenced on an Illumina sequencer.

The three clusters consist of:

* Cluster 1
  + Patient A1 (2011) - sample ERR029207
  + Patient A2 (2011) - sample ERR029206
  + Patient A3 (2008) - sample ERR026478
* Cluster 2
  + Patient B1 (2001) - sample ERR026474
  + Patient B2 (2012) - sample ERR026473
* Cluster 3
  + Patient C1 (2011) - sample ERR026481
  + Patient C2 (2016) - sample ERR026482

The second goal of this lab is to affirm or dispute the inferred transmissions by comparing the bacterial genomes with each other. We will do that by identifying SNPs between the genomes.

## Getting your project started

Project organization is one of the most important parts of a sequencing project, and yet is often overlooked amidst the excitement of getting a first look at new data. Of course, while it's best to get yourself organized before you even begin your analyses,it's never too late to start, either.

Genomics projects can quickly accumulate hundreds of files across tens of folders. Every computational analysis you perform over the course of your project is going to create many files, which can especially become a problem when you'll inevitably want to run some of those analyses again. For instance, you might have made significant headway into your project, but then have to remember the PCR conditions you used to create your sequencing library months prior.

Other questions might arise along the way:

-   What were your best alignment results?
-   Which folder were they in: Analysis1, AnalysisRedone, or AnalysisRedone2?
-   Which quality cutoff did you use?
-   What version of a given program did you implement your analysis in?

In this exercise we will setup a file system for the project we will be working on during this workshop.

We will start by creating a sub directory that we can use for the rest of the labs. First navigate to your own directory on Uppmax and start an `interactive` session. Then confirm that you are in the correct directory using the `pwd` command.

```bash
cd /proj/g2020004/nobackup/3MK013/<username>
interactive -A uppmax2024-2-10 -M snowy -t 04:00:00
pwd
```

You should see the output:

```output
/proj/g2020004/nobackup/3MK013/<username>
```

::: challenge
## Exercise

Use the `mkdir` command to make the following directories:

```         
molepi
molepi/docs  
molepi/data  
molepi/results
```

::: solution
## Solution

```bash
mkdir molepi
mkdir molepi/docs
mkdir molepi/data
mkdir molepi/results
```
:::
:::

Use `ls -R` to verify that you have created these directories. The `-R` option for `ls` stands for recursive. This option causes `ls` to return the contents of each subdirectory within the directory iteratively.

```bash
ls -R molepi
```

You should see the following output:

```output
molepi/:
data  docs  results

molepi/data:

molepi/docs:

molepi/results: 
```

## Selection of a reference genome

Reference sequences (including many pathogen genomes) are available at [NCBI's refseq database](https://www.ncbi.nlm.nih.gov/refseq/)

A reference genome is a genome that was previously sequenced and is closely related to the isolates we would like to analyse. The selection of a closely related reference genome is not trivial and will warrant an analysis in itself. However, for simplicity, here we will work with the *M. tuberculosis* reference genome H37Rv.

### Download reference genomes from NCBI

Download the *M.tuberculosis* reference genome from the NCBI ftp site.

First, we switch to the data folder to store all our data

```bash
cd molepi/data 
```

The reference genome will be downloaded programmatically from NCBI with `wget`. `wget` is a computer program that retrieves content from web servers. Its name derives from World Wide Web and get.

```bash
wget ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/195/955/GCF_000195955.2_ASM19595v2/GCF_000195955.2_ASM19595v2_genomic.fna.gz
```

This file is compressed as indicated by the extension of `.gz`. It means that this file has been compressed using the `gzip` command.

Extract the file by typing

```bash
gunzip GCF_000195955.2_ASM19595v2_genomic.fna.gz
```

Make sure that is was extracted

```bash
ls
```

``` output
GCF_000195955.2_ASM19595v2_genomic.fna
```

::: challenge
## Challenge: What is the size of the genome?

Find out how many nucleotides the genome has. Hint:

```bash   
module load bioinfo-tools SeqKit
seqkit -h
```

Get assembly statistics from FASTA and FASTQ files.

::: solution
## Solution

The genome has 4'411'532 bp.

``` bash
seqkit stats GCF_000195955.2_ASM19595v2_genomic.fna
```

``` output
file                                    format  type  num_seqs    sum_len    min_len    avg_len    max_len
GCF_000195955.2_ASM19595v2_genomic.fna  FASTA   DNA          1  4,411,532  4,411,532  4,411,532  4,411,532

```

::::::::::::

::::::::::::

## Loops

*Loops* are key to productivity improvements through automation as they allow us to execute commands repeatedly. Similar to wildcards and tab completion, using loops also reduces the amount of typing (and typing mistakes). Our next task is to download our data (see the introduction to this episode) from the short read archive [(SRA) at the European Nucleotide Archive (ENA)](https://www.ebi.ac.uk/ena). There are many repositories for public data. Some model organisms or fields have specific databases, and there are ones for particular types of data. Two of the most comprehensive are the National Center for Biotechnology Information (NCBI) and European Nucleotide Archive (EMBL-EBI). In this lesson we're working with the ENA, but the general process is the same for any database.

We can do this one by one but given that each download takes about one to two hours, this could keep us up all night. Instead of downloading one by one we can apply a loop. Let's see what that looks like and then we'll discuss what we're doing with each line of our loop.

```bash
for filename in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  wget --spider ftp://ftp.sra.ebi.ac.uk/vol1/fastq/"${filename:0:6}"/"${filename}"/"${filename}"_*.fastq.gz
done
```

When the shell sees the keyword `for`, it knows to repeat a command (or group of commands) once for each item in a list. Each time the loop runs (called an iteration), an item in the list is assigned in sequence to the **variable**, and the commands inside the loop are executed, before moving on to the next item in the list.

Inside the loop, we call for the variable's value by putting `$` in front of it. The `$` tells the shell interpreter to treat the **variable** as a variable name and substitute its value in its place, rather than treat it as text or an external command.

In this example, the list is seven accession numbers of reads belonging to the genomes we are interested in. Each time the loop iterates, it will assign a file name to the variable `filename` and run the `wget` command. The first time through the loop, `$filename` is `ERR029207`. The FTP site at EBI is constructed so that fastq files are stored in a `vol1/fastq` subfolder. Read files are not directly in that subfolder, but grouped in subfolders starting with the six first characters of the accession number (`ERR029` in the first case), then in the accession number subfolder. To get to the right place, the file path is constructed with the root path (`ftp://ftp.sra.ebi.ac.uk/vol1/fastq/`), to which the first 6 characters of the accession number is added through that special variable operation (`"${filename:0:6}"`) to get to the right subfolder. Finally, the correct accession subfolder (`"${filename}"`), the file name prefix  (`"${filename}"_*.fastq.gz"`) and the suffix (`_*.fastq.gz`) are added. The `*` ensures that we get reads from both ends.

Use `{}` to wrap the variable so that .fastq.gz will not be interpreted as part of the variable name. In addition, quoting the shell variables is a good practice AND necessary if your variables have spaces in them.

For the second iteration, `$filename` becomes `ERR029206`.

We added a `--spider` option to `wget` just for this exercise, to retrieve only the names of the files and not the actual file, to spare some time downloading files. The data is present in the `data/fastq` subfolder of our group folder:

```bash
ls /proj/g2020004/nobackup/3MK013/data/fastq
```

```output
ERR026473_1.fastq.gz  ERR026474_2.fastq.gz  ERR026481_1.fastq.gz  ERR026482_2.fastq.gz	ERR029207_1.fastq.gz
ERR026473_2.fastq.gz  ERR026478_1.fastq.gz  ERR026481_2.fastq.gz  ERR029206_1.fastq.gz	ERR029207_2.fastq.gz
ERR026474_1.fastq.gz  ERR026478_2.fastq.gz  ERR026482_1.fastq.gz  ERR029206_2.fastq.gz
```

For more, check [Bash Pitfalls](https://mywiki.wooledge.org/BashPitfalls)

::: callout
## Follow the Prompt

The shell prompt changes from `$` to `>` and back again as we were typing in our loop. The second prompt, `>`, is different to remind us that we haven't finished typing a complete command yet. A semicolon, `;`, can be used to separate two commands written on a single line.
:::

::: callout
## Same Symbols, Different Meanings

Here we see `>` being used a shell prompt, whereas `>` is also used to redirect output. Similarly, `$` is used as a shell prompt, but, as we saw earlier, it is also used to ask the shell to get the value of a variable.

If the *shell* prints `>` or `$` then it expects you to type something, and the symbol is a prompt.

If *you* type `>` or `$` yourself, it is an instruction from you that the shell to redirect output or get the value of a variable.
:::

We have called the variable in this loop `filename` in order to make its purpose clearer to human readers. The shell itself doesn't care what the variable is called; if we wrote this loop as:

```bash
for x in ERR01 ERR02 ERR03
do
  echo ftp://ftp.sra.ebi.ac.uk/"${x}".fastq.gz
done
```

or:

```bash
for pizza in ERR01 ERR02 ERR03
do
  echo ftp://ftp.sra.ebi.ac.uk/"${pizza}".fastq.gz
done
```

it would work exactly the same way. *Don't do this.* Programs are only useful if people can understand them, so meaningless names (like `x`) or misleading names (like `pizza`) increase the odds that the program won't do what its readers think it does.

::: callout
## Multipart commands

The `for` loop is interpreted as a multipart command. If you press the up arrow on your keyboard to recall the command, it may be shown like so (depends on your bash version):

```bash
for filename in ERR01 ERR02 ERR03; do echo ftp://ftp.sra.ebi.ac.uk/"${filename}".fastq.gz ; done
```

When you check your history later, it will help your remember what you did!
:::

All the read files start with ERR and have a `fastq.gz` extension. It means they are in fastq format and compressed. Softlink the files in your `molepi/data` folder, using a nested loop. The `data/fastq` folder is three levels up:

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/data
for filename in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  for end in 1 2
  do
    ln -s ../../../data/fastq/"${filename}"_"${end}".fastq.gz .
  done
done
```

@@

![](fig/Workflow.png){alt="Schematic workflow of the bioinformatics workflow"}

@@

![](fig/var_calling_workflow_qc.png){alt="Quality control workflow"}

@@

| Line | Description                                                                                                  |
|------------|------------------------------------------------------------|
| 1    | Always begins with '\@' and then information about the read                                                  |
| 2    | The actual DNA sequence                                                                                      |
| 3    | Always begins with a '+' and sometimes the same info in line 1                                               |
| 4    | Has a string of characters which represent the quality scores; must have same number of characters as line 2 |

We can view the first complete read in one of the files our dataset by using `head` to look at the first four lines. You will use zcat to extract text from a gzipped file:

```bash
zcat ERR029206_1.fastq.gz | head -n4 
```

@@

We can now see that the quality of an `N` (`'`) is 6, which corresponds to a over 25% error probability (\$10\^{-\frac{6}{10}} \cong 0.25 \$).

> ### Quality Encodings Vary
>
> Although we've used a particular quality encoding system to demonstrate interpretation of read quality, different sequencing machines use different encoding systems. This means that, depending on which sequencer you use to generate your data, a `#` may not be an indicator of a poor quality base call.

::: callout
This mainly relates to older Solexa/Illumina data, but it's essential that you know which sequencing platform was used to generate your data, so that you can tell your quality control program which encoding to use. If you choose the wrong encoding, you run the risk of throwing away good reads or (even worse) not throwing away bad reads!
:::

@@

![](fig/good_quality1.8.png){alt="A FastQC output showing reads with good quality"}

@@

![](fig/bad_quality1.8.png){alt="A FastQC output showing reads with bad quality"}

@@

``` bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/data
```

::: challenge
### Exercise

We softlinked the samples. How many FASTQ files are in this dataset? Why is this? What is the difference between the two files that start with the same name? Discuss with your neighbour.

::: solution
### Solution

There are 14 FASTQ files. Each sample has [paired-end reads](https://thesequencingcenter.com/knowledge-base/what-are-paired-end-reads/).
:::
:::

FastQC can accept multiple file names as input, including files that are compressed (`.gz`) so we can use the `*.fastq.gz` wildcard to run FastQC on both the compressed FASTQ files for sample ERR029206 in this directory.

```bash
module load FastQC
fastqc ERR029206*.fastq.gz
```

@@

```bash
ls
```

```output
...
[...]
ERR029206_1.fastq.gz
ERR029206_1_fastqc.html
ERR029206_1_fastqc.zip
ERR029206_2.fastq.gz
ERR029206_2_fastqc.html
ERR029206_2_fastqc.zip
[...]
```

@@

If we are working on our local computers, we may display each of these HTML files as a webpage. Otherwise, copy the files to your local computer using `scp` and use your browser:

```bash
firefox ERR029206_1_fastqc.html
```

@@

::: challenge
### Exercise

Discuss your results with a neighbor. Do the sample looks good in terms of per base sequence quality?
:::

::: challenge
### Exercise

What are the read lengths of the different samples?
:::

@@

```bash
module load seqtk
seqtk
```

@@

| option   | meaning                                   |
|----------|-------------------------------------------|
| `seq`    | common transformation of FASTA/Q          |
| `comp`   | get the nucleotide composition of FASTA/Q |
| `trimfq` | trim FASTQ using the Phred algorithm      |

@@

```bash
seqtk trimfq
```

| step | meaning                                                 |
|------|---------------------------------------------------------|
| `-q` | error rate threshold (disabled by -b/-e) [0.05]         |
| `-l` | maximally trim down to INT bp (disabled by -b/-e) [30]  |
| `-b` | trim INT bp from left (non-zero to disable -q/-l) [0]   |
| `-e` | trim INT bp from right (non-zero to disable -q/-l) [0]. |

@@

```bash
seqtk trimfq -q 0.01 ERR01_1.fastq.gz > ERR01_1_trim.fastq
```

@@

```bash
cd molepi/data/
```

@@

```bash
seqtk trimfq -q 0.01 ERR029206_1.fastq.gz > ERR029206_1_trim.fastq
```

@@

::: challenge
### Exercise

@@

::: solution
### Solution

```bash
seqtk fqchk ERR029206_1.fastq.gz | head -n 3
```

@@

```output
min_len: 30; max_len: 76; avg_len: 69.12; 36 distinct quality values
POS #bases  %A  %C  %G  %T  %N  avgQ    errQ    %low    %high
ALL 287211920   18.2    32.0    31.9    17.8    0.0 31.7    28.8    1.3 98.7
```
:::
:::

@@

```bash
for infile in *.fastq.gz
do
  outfile="${infile}"_trim.fastq
  seqtk trimfq -q 0.01 "${infile}" >  "${outfile}" 
done
```

@@

``` bash
ls
```

``` output
...
ERR029206_1.copy.fastq 
ERR029206_1.fastq.gz
ERR029206_1.fastq.gz_trim.fastq 
ERR029206_1_trim.fastq 
ERR029206_2.fastq.gz
ERR029206_2.fastq.gz_trim.fastq
GCF_000195955.2_ASM19595v2_genomic.fna
```

@@

``` bash
mkdir trimmed_fastq
mv *fastq.gz_trim* trimmed_fastq
cd trimmed_fastq
ls
```

``` output
...
ERR029206_1.fastq.gz_trim.fastq 
ERR029206_2.fastq.gz_trim.fastq 
```

::: challenge
### Challenge

@@

::: solution
### Solution

@@

Coverage = #bases (forward + reverse) / genome size In this case: 125.18 = (287211920 + 265048852) / 4411532
:::
:::

::: keypoints

- `wget` is a computer program to get data from the internet
- `for` loops let you perform the same set of operations on multiple files with a single command
- Sequencing data is large
- In bioinformatic workflows the output of one tool is the input of the other.
- FastQC is used to judge the quality of sequencing reads.
- Data cleaning is an essential step in a genomics pipeline.
- We will work towards confirming or disputing transmission in TB cases
- After this practical training you will have some familiarity with working on the command line

:::
